\documentclass[12pt]{amsart}
\usepackage[T1]{fontenc}
\usepackage{amsmath}

\title{Space-time adaptivity for parabolic evolution equations}
\author{Rob Stevenson \and Raymond van Veneti\"e \and Jan Westerdiep$^\star$}
\address{$^\star$ Korteweg-de Vries Institute for Mathematics, University of Amsterdam}
\email[Jan Westerdiep]{j.h.westerdiep@uva.nl}

\begin{document}

\maketitle

\section*{ABSTRACT}


Taking the well-posed mixed simultaneous space-time variational formulation
introduced in~\cite{And13}, we use methods previously developed in~\cite{SW20}
to construct an adaptive loop that produces space-time approximations as linear
combinations of tensor-products of wavelets in time and finite elements in space.

Using an efficient and reliable `hierarchical basis' error estimator, we apply bulk
chasing to show convergence of the iterands. Moreover, we provide an algorithm for
linear-complexity application of the system matrix and an optimal preconditioner.
Lastly, we include an extensive numerical study to show that this method is competitive
in terms of speed and moreover exhibits optimal convergence rate in the number of
degrees of freedom.

\begin{thebibliography}{99}

\bibitem[And13]{And13}
R.~Andreev.
\newblock Stability of sparse space-time finite element discretizations of
 linear parabolic evolution equations.
\newblock {\em IMA J. Numer. Anal.}, 33(1):242--260, 2013.

\bibitem[SW20]{SW20}
\newblock Stability of Galerkin discretizations of a mixed space-time variational
   formulation of parabolic evolution equations.
\newblock {\em IMA J. Numer. Anal.}, 2020.

\end{thebibliography}


\end{document}
